---
title: "results"
output: word_document
date: "2023-02-20"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

```{r packages}
library(tidyverse);library(readxl);library(lme4);library(emmeans);library(dplyr);library(cowplot);library(knitr);library(broom);library(rmarkdown);library(tinytex);library(tidyr);library(arsenal);library(magrittr);library(dabestr);library(nlme); library(pbkrtest);library(ggpubr); library(nlme)
```

## Results

```{r glucose data + model}
## Scripts:
# - gluc.responsechange.R


# Data
glu.change <- readRDS("./data/data-gen/glucose/glu.change.RDS") # Untransformed data, used in figures
glu.logchange <- readRDS("./data/data-gen/glucose/glu.logchange.RDS") # Log-transformed data, used in the analysis

## Model
# Untransformed
glu.mod <- lmerTest::lmer(change ~ 0 + baseline + time + supplement:time + (1|subject),
                     data = glu.change)
glu.res <- plot(glu.mod)
glu.sum <- summary(glu.mod)

# Log-transformed
glu.lmod <- lmerTest::lmer(change ~ 0 + baseline + time + supplement:time + (1|subject),
                     data = glu.logchange)
glu.lres <- plot(glu.lmod)
glu.lsum <- summary(glu.lmod)

## Emmeans
# Untransformed
glu.emm <- confint(emmeans(glu.mod, specs = ~"supplement|time")) %>%
        data.frame()

# Log-transformed
glu.lemm <- confint(emmeans(glu.lmod, specs = ~"supplement|time")) %>%
        data.frame()

# For inline plotting
bcoef <- coef(summary(glu.lmod))
pscient.120 <- round(bcoef[8,5], 30)
p.120 <- sprintf("%.5f", 1.700971e-16)
pscient.135 <- round(bcoef[9,5], 30)
p.135 <- sprintf("%.5f", 4.299568e-11)
pscient.150 <- round(bcoef[10,5], 30)
p.150 <- sprintf("%.5f", 1.445712e-11)
p.270 <- round(bcoef[11,5], 2)
```

```{r c-peptide data + model}
## Scripts
# - insulin.change.R

## Data
cpep.change <- readRDS("./data/data-gen/glucose/cpep.change.RDS") # Untransformed data, for figures
cpep.lchange <- readRDS("./data/data-gen/glucose/cpep.logchange.RDS") # Log-transformed data, for analysis


## Linear mixed effects model
# Untransformed
cpep.mod <- lmerTest::lmer(change ~ 0 + baseline + time + supplement:time + (1|subject),
                     data = cpep.change) 
cpep.res <- plot(cpep.mod)
cpep.sum <- summary(cpep.mod)

# Log-transformed
cpep.lmod <- lmerTest::lmer(change ~ 0 + baseline + time + supplement:time + (1|subject),
                     data = cpep.lchange) 
cpep.lres <- plot(cpep.lmod)
cpep.lsum <- summary(cpep.lmod)

## Emmeans
# Untransformed
cpep.emm <- confint(emmeans(cpep.mod, specs = ~"supplement|time")) %>%
  data.frame()

# Log-transformed
cpep.lemm <- confint(emmeans(cpep.lmod, specs = ~"supplement|time")) %>%
  data.frame()

# For inline plotting
cpcoef <- coef(summary(cpep.lmod))
pscient.120 <- round(cpcoef[6,5], 30)
cpp.120 <- sprintf("%.5f", 3.229267e-15)
cpscient.150 <- round(cpcoef[7,5], 30)
cpp.150 <- sprintf("%.5f", 1.414491e-15)
cpp.270 <- round(cpcoef[8,5], 2)

```

```{r strength model}
## Scripts:
# - Strength.index.R

## Data
# Full data - baseline -> 23h post 6RT
str.change <- readRDS("./data/data-gen/humac/str.change.RDS") # Untransfromed data
str.lchange <- readRDS("./data/data-gen/humac/str.lchange.RDS") # Log-transfromed data

# Acute data - Post 5RT -> 23h post 6RT
astr.change <- readRDS("./data/data-gen/humac/astr.change.RDS") # Untransfromed data
astr.lchange <- readRDS("./data/data-gen/humac/astr.lchange.RDS") # Log-transfromed data


## Linear mixed effects model full data

# Untransformed model
str.mod <- lmerTest::lmer(change ~ 0 + baseline + time + supplement:time + (1|subject),
                         data = str.change)
str.res <- plot(str.mod)
str.sum <- summary(str.mod)

# Log-transformed model
str.lmod <- lmerTest::lmer(change ~ 0 + baseline + time + supplement:time + (1|subject),
                        data = str.lchange)
str.lres <- plot(str.lmod)
str.lsum <- summary(str.lmod)

## Emmeans full data
# Untransformed
str.emm <- confint(emmeans(str.mod, specs = ~"supplement|time")) %>%
  data.frame() 

# Log-transformed
str.lemm <- confint(emmeans(str.lmod, specs = ~"supplement|time")) %>%
  data.frame() 

## Linear mixed effects model acute
# Untransformed
astr.mod <- lmerTest::lmer(change ~ 0 + baseline + time + supplement:time + (1|subject),
                        data = astr.change)

astr.res <- plot(astr.mod)
astr.sum <- summary(astr.mod)

# Log-transformed
astr.lmod <- lmerTest::lmer(change ~ 0 + baseline + time + supplement:time + (1|subject),
                         data = astr.lchange)

astr.lres <- plot(astr.lmod)
astr.lsum <- summary(astr.lmod)

## Emmeans acute data
astr.emm <- confint(emmeans(astr.mod, specs = ~"supplement|time")) %>%
  data.frame() 

astr.lemm <- confint(emmeans(astr.lmod, specs = ~"supplement|time")) %>%
  data.frame() 


## For inline plotting full data
strcoef <- coef(summary(str.mod))

# Time points compared to baseline (full data)
pc2 <- round(strcoef[2,5], 3)
pc3 <- round(strcoef[3,5], 3)
pscient.c4 <- round(strcoef[4,5], 40)
pc4 <- sprintf("%.5f", 7.886277e-06)
pc5 <- round(strcoef[5,5], 3)
pc6 <- round(strcoef[6,5], 3)
pc7 <- round(strcoef[7,5], 5)


# Glucose compared to placebo (full data)
pdiff.c2 <- round(strcoef[8,5], 3)
pdiff.c3 <- round(strcoef[9,5], 3)
pdiff.c4 <- round(strcoef[10,5], 3)
pdiff.c5 <- round(strcoef[11,5], 3)
pdiff.c6 <- round(strcoef[12,5], 3)
pdiff.c7 <- round(strcoef[13,5], 3)

## For inline plotting acute data
astrcoef <- coef(summary(astr.mod))
apc2 <- round(astrcoef[2,5], 3)
apc3 <- round(astrcoef[3,5], 3)
apc4 <- round(astrcoef[4,5], 3)




```

```{r nutrition model}
# Scripts
# - nut.change.R

# Data
pro.change <- readRDS("./data/data-gen/nutrition/pro.change.RDS")
fat.change <- readRDS("./data/data-gen/nutrition/fat.change.RDS")
cho.change <- readRDS("./data/data-gen/nutrition/cho.change.RDS")
cal.change <- readRDS("./data/data-gen/nutrition/cal.change.RDS")
propr.change <- readRDS("./data/data-gen/nutrition/propr.change.RDS")

## Models
# Protein
pro.mod <- lmerTest::lmer(change ~ 0 + d1 + time + supplement:time + (1|subject),
                     data = pro.change)

pro.res <- plot(pro.mod)
pro.sum <- summary(pro.mod)


# Fat
fat.mod <- lmerTest::lmer(change ~ 0 + d1 + time + supplement:time + (1|subject),
                     data = fat.change)

fat.res <- plot(fat.mod)
fat.sum <- summary(fat.mod)


# Carbohydrates
cho.mod <- lmerTest::lmer(change ~ 0 + d1 + time + supplement:time + (1|subject),
                     data = cho.change)

cho.res <- plot(cho.mod)
cho.sum <- summary(cho.mod)


# Calories
cal.mod <- lmerTest::lmer(change ~ 0 + d1 + time + supplement:time + (1|subject),
                     data = cal.change)

cal.res <- plot(cal.mod)
cal.sum <- summary(cal.mod)


# Protein/kg
propr.mod <- lmerTest::lmer(change ~ 0 + d1 + time + supplement:time + (1|subject),
                     data = propr.change)

propr.res <- plot(propr.mod)
propr.sum <- summary(propr.mod)

## Emmeans

# Protein
protein.emm <- confint(emmeans(pro.mod, specs = ~"supplement|time")) %>%
  data.frame()

# Fat
fat.emm <- confint(emmeans(fat.mod, specs = ~"supplement|time")) %>%
  data.frame()

# Carbohydrates
cho.emm <- confint(emmeans(cho.mod, specs = ~"supplement|time")) %>%
  data.frame()

# Calories
cal.emm <- confint(emmeans(cal.mod, specs = ~"supplement|time")) %>%
  data.frame()

# Protein/kg
propr.emm <- confint(emmeans(propr.mod, specs = ~"supplement|time")) %>%
  data.frame()


```

```{r volume model}
## Scripts
# - volume.cleanup.R
# - volume.analysis.R

## Data
vol.change <- readRDS("./data/data-gen/training/vol.change.RDS") # Untransformed data
vol.lchange <- readRDS("./data/data-gen/training/vol.lchange.RDS") # Log-transformed data

## Models

# Untransformed 
vol.mod <- lmerTest::lmer(change ~ 0 + baseline + time + supplement:time + (1|subject),
                     data = vol.change)
vol.res <- plot(vol.mod)
vol.sum <- summary(vol.mod)

# Log-transformed
vol.lmod <- lmerTest::lmer(change ~ 0 + baseline + time + supplement:time + (1|subject),
                     data = vol.lchange)

vol.lres <- plot(vol.lmod)
vol.lsum <- summary(vol.lmod)

## Emmeans

# Untransformed
vol.emm <- confint(emmeans(vol.mod, specs = ~"supplement|time")) %>%
  data.frame()

# Log-transformed
vol.lemm <- confint(emmeans(vol.lmod, specs = ~"supplement|time")) %>%
  data.frame()

## For inline plotting

# Sessions compared to baseline
volcoef <- coef(summary(vol.mod))
pvol.s2 <- round(volcoef[2,5], 4)
pvol.s3 <- round(volcoef[3,5], 4)
pvol.s4 <- round(volcoef[4,5], 4)
pvol.s5 <- round(volcoef[5,5], 3)
pvol.s6 <- round(volcoef[6,5], 5)

# Glucose compared to plac
pvol.c2 <- round(volcoef[7,5], 3) 
pvol.c3 <- round(volcoef[8,5], 3) 
pvol.c4 <- round(volcoef[9,5], 3) 
pvol.c5 <- round(volcoef[10,5], 3) 
pvol.c6 <- round(volcoef[11,5], 3) 


```


There was no difference in mean change of total training session volume between exercising with placebo or glucose, with an increase of 18 and 16% from session 1 to session 6, respectively (Tab 2; Mean change, 2: *p* = `r pvol.c2`, 3: *p* = `r pvol.c3`, 4: *p* = `r pvol.c4`, 5: *p* = `r pvol.c5`, 6: *p* = `r pvol.c6`). Both exercising with glucose and placebo led to significant increases in mean total training session volume at session four, five and six, compared to session 1 (Tab 2; Mean change, 2: *p* = `r pvol.s2`, 3: *p* = `r pvol.s3`, 4: *p* = `r pvol.s4`, 5: *p* = `r pvol.s5`, 6: *p* = `r pvol.s6`). In regard to the dietary intervention, there were no difference in mean macro nutrient intake (protein, fat, carbohydrates) or total calorie intake between glucose and placebo on pairwise consecutive days (*p* > 0.05). 

Glucose ingestion before and after RT led to significant increases in plasma levels of glucose by 38% immediately before RT (Fig 1B, 0 min: *p* = `r p.120`), by 31% during RT (1B, 15 min: *p* = `r p.135`) and by 32% immediately after RT (1B, 30 min: *p* = `r p.150`), compared to placebo ingestion before and after RT. Two hours after the session, RT with glucose measured 8% lower plasma glucose levels compared to RT with placebo (1B, 270min: *p* = `r p.270`). Similarly, levels of c-peptide increased significantly more in exercises with glucose compared to placebo, by 95% immediately before (Fig 1C, 0 min: *p* = `r cpp.120`) and 87% after RT (1C, 30 min: *p* = `r cpp.150`). 

After 5 RT sessions (Fig 1D, Post 5RT), there was a significant difference in strength between glucose and placebo,  where placebo decreased strength 7% more than glucose (1D, *p* = `r pdiff.c4`). At the remaining time points, there were no significant differences in mean change between exercising with glucose and placebo (1D, Post 2RT: *p* = `r pdiff.c2`, Post 4RT: *p* = `r pdiff.c3`, 30min post 6RT: *p* = `r pdiff.c5`, 2h post 6RT: *p* = `r pdiff.c6`, 23h post 6RT: *p* = `r pdiff.c7`). Both RT with glucose and placebo led to significantly reduced strength post fifth session compared to baseline, by 11 and 18% respectively (1D, Post 5RT: *p* = `r pc4`). Comparisons of the acute data gathered from post fifth session until and including 23hrs post sixth session showed a significant mean increase in strength of 5-9% after RT with glucose and placebo 30min after sixth RT session (1D, 30min post 6RT: *p* = `r apc2`) and two hours after sixth RT session (1D, 2h post 6RT: *p* = `r apc3`). Twenty-three hours after the last RT, strength was unchanged compared to after fifth RT session (23h post 6RT: *p* = `r apc4`).

### Markers of ribosome biogenesis

#### Total RNA and ribosomal RNA

```{r total RNA model}
## Scripts
# - totrna.cleanup.R
# - totrna.analysis.R


# Data
trna.lchange <- readRDS("./data/data-gen/rna/totrna.change.RDS")

# Create model: 
trna.lmod <- lmerTest::lmer(change ~ pre + supplement + (1|subject), 
           data = trna.lchange)
    
trna.lres <- plot(trna.lmod)
trna.lsum <- summary(trna.lmod)

### Get estimated means from the model
trna.emm <- confint(emmeans(trna.lmod, specs = ~"supplement")) %>%
    data.frame()

# For inline reporting
trcoef <- coef(summary(trna.lmod))
trdiff <- round(trcoef[3,1], 3)
trdiffp <- round(trcoef[3,5], 3)
trdiffse <- round(trcoef[3,2], 2)

```

```{r rRNA data model}
## Scripts
# - qpcr-import
# - qpcr-analysis
# - rrna.change
# - rrnachange.figs

# Log-transformed data
lchange.18 <- readRDS("./data/data-gen/rna/lchange.18.RDS")
lchange.28 <- readRDS("./data/data-gen/rna/lchange.28.RDS")
lchange.58 <- readRDS("./data/data-gen/rna/lchange.58.RDS")
lchange.5 <- readRDS("./data/data-gen/rna/lchange.5.RDS")
lchange.47 <- readRDS("./data/data-gen/rna/lchange.47.RDS")

# Create model: 
# Needs to have an intercept per participant (mixed model)
# Control for pre values.

# 18S
lmod.18 <- lmerTest::lmer(change ~ pre + supplement + (1|subject), 
                     data = lchange.18)
lres.18 <- plot(lmod.18)
lsum.18 <- summary(lmod.18)

# 28S
lmod.28 <- lmerTest::lmer(change ~ pre + supplement + (1|subject), 
                     data = lchange.28)
lres.28 <- plot(lmod.28)
lsum.28 <- summary(lmod.28)

# 5.8S
lmod.58 <- lmerTest::lmer(change ~ pre + supplement + (1|subject), 
                     data = lchange.58)
lres.58 <- plot(lmod.58)
lsum.58 <- summary(lmod.58)

# 5S
lmod.5 <- lmerTest::lmer(change ~ pre + supplement + (1|subject), 
                     data = lchange.5)
lres.5 <- plot(lmod.5)
lsum.5 <- summary(lmod.5)

# 47S
lmod.47 <- lmerTest::lmer(change ~ pre + supplement + (1|subject), 
                     data = lchange.47)
lres.47 <- plot(lmod.47)
lsum.47 <- summary(lmod.47)

### Get estimated means from the model, these are average increase at 
# pre = 0 (the average pre value)
# remember that these are originally log-fold change values, transformed back using exp(emmean) etc

# 18S
lemm.18 <- confint(emmeans(lmod.18, specs = ~"supplement")) %>%
        data.frame()

# 28S
lemm.28 <- confint(emmeans(lmod.28, specs = ~"supplement")) %>%
        data.frame()

# 5.8S
lemm.58 <- confint(emmeans(lmod.58, specs = ~"supplement")) %>%
        data.frame()

# 5S
lemm.5 <- confint(emmeans(lmod.5, specs = ~"supplement")) %>%
        data.frame()

# 47S
lemm.47 <- confint(emmeans(lmod.47, specs = ~"supplement")) %>%
        data.frame()

## For inline reporting

# 18S
rr18coef <- coef(summary(lmod.18))
rr18diffp <- round(rr18coef[3,5], 3)
rr18diffse <- round(rr18coef[3,2], 3)

# 28S
rr28coef <- coef(summary(lmod.28))
rr28diffp <- round(rr28coef[3,5], 3)
rr28diffse <- round(rr28coef[3,2], 3)

# 5.8S
rr58coef <- coef(summary(lmod.58))
rr58diffp <- round(rr58coef[3,5], 3)
rr58diffse <- round(rr58coef[3,2], 3)

# 5S
rr5coef <- coef(summary(lmod.5))
rr5diffp <- round(rr5coef[3,5], 3)
rr5diffse <- round(rr5coef[3,2], 3)

# 47S
rr47coef <- coef(summary(lmod.47))
rr47diffp <- round(rr47coef[3,5], 3)
rr47tdiffp <- round(rr47coef[1,5], 3)
rr47diffse <- round(rr47coef[3,2], 3)

```

Resistance training with glucose did not induce a higher accumulation of total RNA (fig 2A, *p* = `r trdiffp`) or rRNA  (fig 2B, 47S: *p* = `r rr47diffp`, 18S: *p* = `r rr18diffp`, 28S: *p* = `r rr28diffp`, 5.8S: *p* = `r rr58diffp`, 5S: *p* = `r rr5diffp`) compared to RT with placebo. From baseline to post-intervention, there was a mean increase in levels of total RNA by 26 and 22% after RT with glucose and placebo, respectively (*p* < 0.05, compared to baseline). A robust accumulation was also observed in ribosomal RNA expression in both RT with glucose and placebo, with mean increases between 34-43% (GLU) and 33-41% (PLAC) (*p* < 0.05, compared to baseline) in the four rRNA’s. The expression of the 47S pre-rRNA increased by 37 and 59% with glucose and placebo respectively, where only RT with placebo increased significantly from baseline to post (fig 45, *p* = `r rr47tdiffp`).

#### Protein

```{r western model}
## Script: 
# - western.import.R
# - western.cleanup.R
# - western.analysis.R

# Data
prot.dat <- readRDS("./data/data-gen/protein/prot.datc.RDS") # from the script: western.cleanup.R


### Modelling target signal per supplement 
# These models try to explain the normalised signal by time, supplement and time+supplement (?). The random = list command creates an individual 
# intercept for the included (e.g., subject, gel.sample). 

## c-Myc model

# Untransformed
cmyc.mod <- lme(norm.sign ~ time + supplement + time:supplement, 
                random = list(subject = ~ 1),
                data = filter(prot.dat, target == "cmyc"))

cmyc.sum <- summary(cmyc.mod)
cmyc.res <- plot(cmyc.mod, resid(., type = "p") ~ fitted(.))



# Log-transformed
cmyc.lmod <- lme(log(norm.sign) ~ time + supplement + time:supplement, 
                 random = list(subject = ~ 1),
                 data = filter(prot.dat, target == "cmyc"))

cmyc.lsum <- summary(cmyc.lmod)
cmyc.lres <- plot(cmyc.lmod, resid(., type = "p") ~ fitted(.))




## UBF model
# Untransformed
ubf.mod <- lme(norm.sign ~ time + supplement + time:supplement, 
               random = list(subject = ~ 1),
               data = filter(prot.dat, target == "ubf"))


ubf.sum <- summary(ubf.mod)
ubf.res <- plot(ubf.mod, resid(., type = "p") ~ fitted(.))

# Log-transformed
ubf.lmod <- lme(log(norm.sign) ~ time + supplement + time:supplement, 
                random = list(subject = ~ 1),
                data = filter(prot.dat, target == "ubf"))


ubf.lsum <- summary(ubf.lmod)
ubf.lres <- plot(ubf.lmod, resid(., type = "p") ~ fitted(.))


# RPS6 model
# Untransformed
rps6.mod <- lme(norm.sign ~ time + supplement + time:supplement, 
                 random = list(subject = ~ 1),
                 data = filter(prot.dat, target == "rps6"))

rps6.sum <- summary(rps6.mod)
rps6.res <- plot(rps6.mod, resid(., type = "p") ~ fitted(.))


# Log-transformed
rps6.lmod <- lme(log(norm.sign) ~ time + supplement + time:supplement, 
          random = list(subject = ~ 1),
          data = filter(prot.dat, target == "rps6"))

rps6.lsum <- summary(rps6.lmod)
rps6.lres <- plot(rps6.lmod, resid(., type = "p") ~ fitted(.))


### Getting emmeans from the models

## c-Myc emmeans
# Untransformed
cmyc.emm <- emmeans(cmyc.mod, specs = ~ time|supplement) %>%
  data.frame()

# Log-transformed
cmyc.lemm <- emmeans(cmyc.lmod, specs = ~ time|supplement) %>%
  data.frame()



## UBF emmeans
# Untransformed
ubf.emm <- emmeans(ubf.mod, specs = ~ time|supplement) %>%
  data.frame()

# Log-transformed
ubf.lemm <- emmeans(ubf.lmod, specs = ~ time|supplement) %>%
  data.frame()



## RPS6
# Untransformed
rps6.emm <- emmeans(rps6.mod, specs = ~ time|supplement) %>%
  data.frame()

# Log-transformed
rps6.lemm <- emmeans(rps6.lmod, specs = ~ time|supplement) %>%
  data.frame()



### For inline reporting
## Diff cmyc

ccoef <- coef(summary(cmyc.lmod))
cdiff <- round(ccoef[4,5], 3)
cpla <- round(ccoef[2,5], 20)

## Diff UBF
ucoef <- coef(summary(ubf.lmod))
udiff <- round(ucoef[4,5], 3)

## Diff rps6
rpcoef <- coef(summary(rps6.lmod))
rpdiff <- round(rpcoef[4,5], 3)

```

```{r correlation model}
# Data 
west.dat <- readRDS("./data/data-gen/protein/prot.datc.RDS") # Quantified and cleaned western data,
                                                               # (outliers marked) normalised by pool

rna.dat <- readRDS("./data/data-gen/rna/rna.dat.RDS") # Quantified and cleaned (outliers marked), 
                                                      # total RNA data, normalised by wet muscle weight


## Prepare data frames for joining

ubf.rdy <- west.dat %>%  # One data frame without outliers
  select(subject, time, outlier, supplement, target, norm.sign) %>%
  filter(target == "ubf",
         outlier == "in") %>%
  select(-target) %>%
  group_by(subject, time, supplement) %>%
  summarise(mean.sign = mean(norm.sign, na.rm = TRUE)) 


rna.rdy <- rna.dat %>%
  select(subject, time, outlier, supplement, RNA.weight) %>%
  filter(outlier == "in") %>%
  mutate(subject = as.character(subject)) %>%
  group_by(subject, time, supplement) %>%
  summarise(mean.rna = mean(RNA.weight, na.rm = TRUE)) 

## Correlation analysis

joined.dat <- ubf.rdy %>%
  right_join(rna.rdy) %>%
  #filter(supplement != "GLUCOSE") %>%
  
  mutate(sd.ubf = (mean.sign - mean(mean.sign, na.rm = TRUE )/sd(mean.sign, na.rm = TRUE))) 


library(nlme)

corr.mod <- lme(mean.rna ~ mean.sign + time , 
         random = list(subject = ~ 1), 
          data = joined.dat, 
         na.action = na.omit)

corr.sum <- summary(corr.mod)
corr.res <- plot(corr.mod)

corr.lmod <- lme(log(mean.rna) ~ log(mean.sign) + time , 
         random = list(subject = ~ 1), 
          data = joined.dat, 
         na.action = na.omit)

corr.lsum <- summary(corr.lmod)
corr.lres <- plot(corr.lmod)

corrcoef <- coef(summary(corr.mod))
corrp <- round(corrcoef[2,5], 3)
```

Levels of c-Myc increased by 51% more after RT with placebo compared to RT with glucose, a significant difference (Fig 3A: *p* = `r cdiff`). Levels of UBF and rpS6 increased by 42 and 33% more in placebo compared to glucose, however, this was not a significant difference (Fig 3A: UBF: *p* = `r udiff`, rpS6: *p* = `r rpdiff`). The increase in both RT with glucose and placebo for all measured proteins was significant at post compared to baseline (*p* < 0.05). Further, we found a linear relationship between UBF and total RNA, where an increase of 1 SD unit of UBF equated to a 13% increase in total RNA (*p* = `r corrp`).

